---
title: "Estimate Centers of Activity for Lake Trout Tagged with an Acoustic Transmitter"
author: Benjamin L. Hlina
date: "2026-06-13"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Estimate Centers of Activity for Lake Trout Tagged with an Acoustic Transmitter}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 7
)
```
## Introduction

This vignette will walk you through the analyses presented in [Winton et al. 2018](https://besjournals.onlinelibrary.wiley.com/doi/full/10.1111/2041-210X.13080), who describes the use of spatial point process models to estimate individual centers of activity (COA) from passive acoustic telemetry data. This vignette walks through how to prepare the data, using the models, and interpretating the results. We will be using the simplest case, which assumes that detection probabilities/receiver detection ranges remain constant over time, to a more complex application of a test-tag integrated model, that incorporates detection data from one or more stationary test transmitters to estimate time-varying detection ranges. The models are fitted in a Bayesian framework using the Stan software ([Carpenter et al. 2017](https://www.jstatsoft.org/article/view/v076i01/0)); code was modified from models found in [Royle et al. 2013](https://www.sciencedirect.com/book/monograph/9780124059399/spatial-capture-recapture) for fitting spatial point process models to data from camera traps. We prefer the Bayesian approach for COA estimation due to the treatment of uncertainty, but realize the longer computational time required may be prohibitive for some applications. We'd also like to note that the models described can support varying degrees of complexity - not all applications will require (or have the data to support) the most complex version of the model. The simpler the model, the shorter the run-time.
       
We have tried to make the instructions outlined in this vignette user-friendly since we are a group of applied biologists with varying degrees of statistical experience. If some of the statistical notation outlined here or in [Winton et al. 2018](https://besjournals.onlinelibrary.wiley.com/doi/full/10.1111/2041-210X.13080) remains unclear, feel free to contact us with questions for clarification. This is a new package, so if you find bugs, places where code efficiency could be improved, or instances where the documentation could be made more user-friendly, please let us know through the [issues]( https://github.com/trackyverse/TelemetrySpace) on the GitHub repository! 

## Models 
 
 For simplicity purposes of this vignette, model specifications can be found in the following documentation: 

- [Standard detection probability](https://telemetryspace.trackyverse.org/articles/coa_standard_gaussian_model.html) (i.e., `COA_standard()`)
- [Time-varying detection probability]() (i.e., `COA_TimeVarying ()`)
- [Tag-integrated time-varying detection probability]() (i.e., `COA_TagInt()`)

 
## Data preparation

To run spatial point process/detection probability models in Stan we need to do a couple of things to our detection and receiver location data. This includes first providing the model with all the necessary information as well as in the proper format. Stan likes to handle data using vectors and arrays, which we often use in R but they are configured differently (e.g., `data.frame()` is a bunch vectors of different classes, while a `matrix` is a bunch of vectors of the same class). These vectors and arrays need to be provided to Stan in a `list`. To learn more about Stan, please click this [link](https://mc-stan.org/). Stan is written in C++, making it quite fast, with Stan programs following a very systematic order, see [link to understand a Stan program]( https://mc-stan.org/docs/reference-manual/blocks.html). Stan uses a [Hamilton Monte Carlo](https://mc-stan.org/docs/2_18/reference-manual/hamiltonian-monte-carlo.html) which is a [Markov chain Monte Carlo (MCMC)](https://en.wikipedia.org/wiki/Markov_chain_Monte_Carlo) that obtains a sequence of random samples whose distribution converges to a target probability distribution that is difficult to sample directly. Stan also employs a  [No U-Turn Sampler (NUTS)](https://mc-stan.org/docs/2_18/stan-users-guide/sampling-difficulties-with-problematic-priors.html) to improve efficiency. We will now walk-through how-to setup the data to be able to run the models. 

We have several example datasets along with several functions to assist and streamline the data preparation. First, we need to evaluate the positions of the receivers and create a [Azimuthal Equal Distance projection (aeqd)](https://en.wikipedia.org/wiki/Azimuthal_equidistant_projection). We use an aeqd project set in kilometers (km) for several reasons, the first being that in Stan it is very efficient to calculate distances among receivers, however, the distance between receivers needs to be on the same x and y plane and ranges need to be between 0.2 - 20 km apart. The second is because....(Mike I need you to add stuff here as you know more about aeqd). We will be working with example data for a single Lake Trout (*Salvelinus namaycush*) that was implanted with an acoustic transmitter in Parry Sound which is a large embayment of Georgian Bay, Lake Huron. 

First, we load all the packages needed to carry out the analysis.
```{r setup, message=FALSE}
{
  library(bayesplot)
  library(ggplot2)
  library(rstan)
  library(sf)
  library(TelemetrySpace)
  library(tidyr)
}
```

### Receiver Locations 

Next, we will look at the location of the receivers in Parry Sound. To build an aqed projection we need the locations to 
be in meters which we can transform these locations to a meter focused coordinate reference system (crs), in this case a projected coordinate system (e.g., UTMs). You will notice that `ps_rec_loc` is a `data.frame`. We need to make it into an `sf` object which is in a geographical coordinate reference system (e.g., WGS 84; degrees longitude and latitude). We need to transform it into a projected coordinate system, in this case we used UTMs. We then can build the aeqd projection. We 
can do this using `st_as_sf()` from [{sf}](https://r-spatial.github.io/sf/). 

```{r recs inspect}
# first look at the example
head(ps_rec_loc)
str(ps_rec_loc)

# make it into a sf object

ps_rec_loc_sf <- ps_rec_loc |>
  st_as_sf(coords = c("deploy_long", "deploy_lat"), crs = 4326)
# next transform it into projected coordinate system (e.g., utm ESPG:32617)

ps_rec_loc_utm <- ps_rec_loc_sf |>
  st_transform(32617)
```

Next, we will build an aeqd projection for this receiver array that we can then transform the locations of the receivers into. 
```{r aeqd}
aeqd_crs <- build_aeqd(ps_rec_loc_utm)
```

We then can transform the receiver locations into our aeqd project and make sure they are in the correct order. 
```{r transform aqed}
ps_rec_loc_aeqd <- ps_rec_loc_sf |>
  st_transform(aeqd_crs) |>
  (\(.) .[order(.$station_no), ])()
```
Now that we have the receiver locations transformed, we need to first index them appropriately, this is because Stan will not be able to handle 
the `station_no` but instead can handle a numerical index value to identify the receivers. 

```{r index rec}
ps_rec_loc_aeqd$rec <- 1:nrow(ps_rec_loc_aeqd)
```

Next, we will transform this into a `data.frame` the contains two columns (i.e.,`vectors`) that we can supply to the Stan model. We also need to build 
the boundary box of the receiver array. The argument `buffer` in `build_bbox()` assumes a 1 km buffer but this can be changed depending on how you would like the boundary box to be created. 
 
```{r build rec}
rec_loc_vec <- build_rec_coords(ps_rec_loc_aeqd)

rec_limits <- build_bbox(rec_loc_vec)
```

### Detection Data 

Now that we have the receiver locations in a format that Stan can handle, we are going to prepare the detection data. First let’s look at the detection data. 
```{r inspect det data}
head(ps_det_example)
str(ps_det_example)
```

We can notice that this detection data consists of 5 columns with 577 rows. To better understand what each field is you can run `?ps_det_example` to review the full documentation.

For our detection data, we have a few things we need to do, the first is we need to build a time bin that we will create COAs. For this data we are going to use 1 hour but this time bin can range from 30 mins - 1 day or more and depends on the questions you are asking and the species you are working with. 

Let's build our time bins

```{r build time bin}
ps_det_example <- build_time_bin(ps_det_example, unit = "1 hour")
head(ps_det_example)
```

You will notice both a `POSIXct` column that is called `time_bin` and a numerical column called `time`. This `time` column is a numerical index of the time bins. 

We have a few more things we need to do to prepare the data, first we need to add in the numerical index of the receiver values that we created in the first section. We can do this by using `merge()` from base or we could use `left_join()` from `{dplyr}` or `merge.data.table()` from `{data.table}`. 
```{r merge rec}
ps_det_example <- merge(
  ps_det_example,
  st_drop_geometry(ps_rec_loc_aeqd),
  by = "station_no"
)
head(ps_det_example)
```
We are starting to get somewhere; you can see that we now have the time and receiver index values lined up. The last two things we need to do is first determine the number of individuals, in this case it will be 1 and the expected number of detections for an individual for each receiver for each time bin, and the number of transmissions this is where `min_delay` and `max_delay` come into play.  

Let’s build the count data and the number of time steps in the data. 

```{r build count}
ps_count_example <- build_counts(
  df = ps_det_example,
  nrec = nrow(ps_rec_loc_aeqd),
  rec_id = ps_rec_loc_aeqd$rec,
  rec_names = ps_rec_loc_aeqd$station_no
)

time_steps <- build_tstep(ps_count_example)
```

Now we can create the number of individuals and the number of transmissions expected within the time bin. 
```{r build nind and trans}
nind <- length(unique(ps_det_example$tag_serial_no))

ntrans <- build_ntrans(ps_det_example)
```

We can finally move on to running the model; however, we need to understand a bit more about Bayesian frameworks before we proceed.

## Bayesian Analysis

We can now estimate the center of activity for lake trout in Parry Sound 
which is a large embayment of Georgian  Bay, Lake Huron. 

There are a few things to know about running a Bayesian analysis, we suggest reading these resources:

1.  [Basics of Bayesian Statistics - Book](https://statswithr.github.io/book/)
2.  [A Student’s Guide to Bayesian Statistics by Ben Lambert](https://sites.math.rutgers.edu/~zeilberg/EM20/Lambert.pdf)
3.  [Statistical rethinking 2 with rstan and the tidyverse by Solomon Kurz](https://solomon.quarto.pub/sr2rstan/)
4.  [Fitting Bayesian Models using Stan and R by Dan Ovando](https://www.weirdfishes.blog/blog/fitting-bayesian-models-with-stan-and-r/)
5.  [Andrew Proctor's - Module 6](https://andrewproctor.github.io/rcourse/module6.html)
6.  [van de Schoot et al., (2021)](https://www.nature.com/articles/s43586-020-00001-2)

### Priors

Bayesian analyses rely on supplying uninformed or informed prior distributions for each parameter (coefficient; predictor) in the model. For the puproses of the deteciton probablity model the following priors are assumed and are not adjustable. To learn more about the structure of the priors and likelihoods see [Standard detection probability](https://telemetryspace.trackyverse.org/articles/coa_standard_gaussian_model.html).

$$
\begin{aligned}
\alpha_0 &\sim \mathrm{Cauchy}(0,\, 2.5), &&\text{truncated to } [-5, 5] \\
\alpha_1 &\sim \mathrm{Cauchy}(0,\, 2.5), &&\text{truncated to } (0, \infty) \\
s_{x,i,t} &\sim \mathrm{Uniform}(x_{\min},\, x_{\max}) &&\text{for all } i, t \\
s_{y,i,t} &\sim \mathrm{Uniform}(y_{\min},\, y_{\max}) &&\text{for all } i, t
\end{aligned}
$$

The Cauchy priors on $\alpha_0$ and $\alpha_1$ are weakly informative and
regularize the intercept and decay-rate parameters toward zero while
allowing heavy tails. Because both parameters carry explicit `lower`/`upper`
bounds in the `parameters` block, the priors are implicitly truncated to
those bounds — Stan automatically renormalizes the density over the
constrained support, so no separate normalizing constant needs to be added
by hand. The activity-center coordinates $s_{x,i,t}$ and $s_{y,i,t}$ have no
explicit sampling statement, so they receive Stan's implicit flat (uniform)
prior over their bounded support $[x_{\min}, x_{\max}]$ and
$[y_{\min}, y_{\max}]$, respectively.


### Model convergence

It is important to always run the model with at least 2 chains. If the model does not converge you can try to increase or decreasing the following:

1.  The number of samples that are discarded; this can be controlled by the argument `warmup`.

2.  The number of iterative samples retained; this can be controlled by the argument `iter`.

3.  The number of samples drawn (in this is controlled by the argument `thin`). 
    This will also reduce the size the posterior distribution kept. 

4.  The `adapt_delta` value using `control = list(adapt_delta = 0.95)`.

When assessing the model, we want $\hat R$ to be 1 or within 0.05 of 1, which indicates that the variance among and within chains are equal (see [{rstan} documentation on $\hat R$](https://mc-stan.org/rstan/reference/Rhat.html)), a high value for effective sample size (ESS), trace plots to look "grassy" or "caterpillar like," and posterior distributions to look relatively normal.

## Model 
We can now run the standard point process/detection probability model. 
```{r run coa_standard}
m <- COA_Standard(
  nind = nind,
  nrec = nrow(ps_rec_loc_aeqd),
  ntime = time_steps,
  ntrans = ntrans,
  y = ps_count_example,
  recX = rec_loc_vec$recX,
  recY = rec_loc_vec$recY,
  xlim = rec_limits$xlim,
  ylim = rec_limits$ylim,
  chains = 2,
  thin = 5
)
```


We can inspect the object created with the first object containing the Stan model. 
To view and work with the model itself call `m$model`. However, as you will see there 
Are a bunch of other objects in the objected created. These objects have pulled important information from the Stan model. 


```{r summary of model object}
summary(m)
```
Let's look at our trace plots for the model parameters and posterior distributions to ensure the model converged properly. Remember the trace plots should look grassy or caterpillar like. A trace plot is the posterior draw for a given iteration plotted with the iteration number on the x axis and the posterior value for a given parameter or latent variable on the y. We evaluate this for both chains and want to see that both chains are converging on a smiler posterior draw for a given parameter or latent variable. We will first look at parameters of the model.  

``` {r trace plots and post dist}
stan_trace(m$model, pars = c("alpha0", "alpha1", "p0", "sigma"))

stan_dens(
  m$model,
  pars = c("alpha0", "alpha1", "p0", "sigma"),
  separate_chains = TRUE,
  linewidth = 0.1
)
```

We can see our trace plots look good and that the posterior distributions of our paramaters look good. 

Next let’s look at our latent variables which are `sx` and `sy` or the estimated locations of the fish based on the detection probability. For large/lengthy time periods it will become cumbersome to evaluate the posteriors 
this way and we recommend inspecting $\hat R$ and ESS. 


``` {r trace plots and post dist latent}
stan_trace(m$model, pars = c("sx[1,1]", "sx[1,2]", "sy[1,1]", "sy[1,2]"))

stan_dens(
  m$model,
  pars = c("sx[1,1]", "sx[1,2]", "sy[1,1]", "sy[1,2]"),
  separate_chains = TRUE,
  linewidth = 0.1
)
```
Again, we can see that everything looks good and our model has converged. 

Now moving on to the other elements in the outputted object from `COA_*()`. 


The second element is a table of parameter estimates, `p0` which is the estimated detection probability at distance of 0 m and $\sigma$ which is units of distance and characterizes the effective spatial range of detection: it is the standard deviation of the Gaussian or logistic decay in detection log-odds with distance from the activity center. A larger $\sigma$ corresponds to a more slow-decaying (longer-range) detection function (`sigma`). The table contains their mean, standard error of the mean, and the associated quantiles from the posterior distributions. The table also includes the effective sample size and the  $\hat R$ statistic (which should be between 0.95 and 1.05).

```{r summary from COA}
m$summary
```

We can see that the mean detection probability at a distance of 0 m is 0.5 or 50% and that the model converged well for these parameters. 

The thrid element returned, is the time required to run the model in minutes. Note that Stan will automatically detect and use multiple cores. If the computer used to run this has multiple cores, the time returned will be longer than the actual run time. This is because it will sum the time for each core. To return the realized run time, divide `m$time` by the number of cores.

```{r runtime }
m$time
```


The fourth element returned, is a summary of the posterior draws. Returned is the `variable`, `median`, and the 2.5 and 97.5% quartiles (e.g., `q2.5` and `q97.5`). This summary `data.frame` provides you the ability to inspect all elements, however, most of the time you will be using this data in a filtered and transformed manner created in other objects in the model output. These objects are more user friendly for plotting and understanding the results. 

```{r summary_draws}
m$summary_draws
```

The fifth element returned, is `data.frame` of the median values for the latent variables `sx` and `sy` which are the estimated location of the fish. The median values can be viewed as the center of activity while the posterior distribution is the overall probability of occurrence around that estimate. The `data.frame` has a few important components `ind` is the index number of the individual, `time` is the index number for a given time bin, `x` and `y` are the median values for a given individual and time, while `x_*` and `y_*` are the 2.5 and 97.5% quantiles for those posteriors, this effectively is viewed as the 95% credible interval, (i.e., Bayesian version of a confidence interval) for each coordinate.

```{r coas}
m$coas
```
The sixth element returned, is a `data.frame` containing the posterior draws from each non-warm-up iteration from all chains. This contains the posterior distribution for each parameter and latent variable for each individual in each time step. It is unlikely that you will use this object instead we have created further objects that organize this data for specific applications that you are more likely to use. 

```{r all draws}
m$all_estimates
```


The seventh element returned is a `data.frame` that contains the extracted and transformed posterior draws for the latent variables `sx` and `sy`. This object we will use to plot our estimates of centers of activity. There are several other columns besides, the draws (i.e, `x`, `y`), the `fish`/ind index number and the `time` index number. We have `.chain`, `.iteration`, `.draw` which identifies the chain, iteration, and draw that posterior draw came from. We also have the column `lp__` which is the unnormalized log posterior density of your model, used for diagnosing sampling efficiency, estimating approximations, and comparing models.

Later on we will plot this object.

```{r loc draws} 
loc_draws <- m$loc_draws
loc_draws
```

The eighth element returned, is a `data.frame` that contains the  contains the extracted and transformed posterior draws for the model parameters `alpha0`, `alpah1`, `p0`, and `sigma`. There are several other columns besides, the draws, the `fish`/ind index number and the `time` index number. We have `.chain`, `.iteration`, `.draw` which identifies the chain, iteration, and draw that posterior draw came from. We also have the column `lp__` which is the unnormalized log posterior density of your model, used for diagnosing sampling efficiency, estimating approximations, and comparing models.  

Later on we will plot this object. 
```{r param draws}
param_draws <- m$param_draws
param_draws
```

The ninth element returned is a `list` that contains generated quantities for `y` from the model. This `y` is different than `sy` and is the estimated counts of detections for each individual at each time step at each receiver. The object `yrep` which is a 2-dimensional array that contains 10 draws (`ndraws` argument controls the number of draws here) and 640 posterior counts of detections the tag number, receiver number and time bin. We will assess to see how well this matches the existing data's distribution as using [posterior probability](https://en.wikipedia.org/wiki/Posterior_probability) which is a type of conditional probability that results from updating the prior probability with information summarized by the likelihood. 

``` {r yrep}
yrep <- m$generated_quantities
yrep
```

## Plotting 

First, we are going to plot the posterior draws for locations. Within the package we 
have a `sf` object that is the shape of Parry Sound. We can use this to plot the posterior draws of `sx` and `sy` to understand the movement of that of lake trout for 8 hours. 

We need to take that `sf` object and transform it into an aeqd projection. 
```{r ps aeqd}
ps_aeqd <- ps |>
  st_transform(aeqd_crs)
```

Now we can plot this within in our system both combined and broken up by time.
```{r plot posteriors}
p <- ggplot() +
  geom_sf(data = ps_aeqd) +
  geom_hex(data = loc_draws, aes(x = x, y = y), bins = 50) +
  scale_fill_viridis_c(option = "H", name = "Posterior\nDensity") +
  scale_x_continuous(n.breaks = 4) +
  theme_bw() +
  theme(
    panel.grid = element_blank(),
    strip.background = element_blank()
  ) +
  labs(
    x = "Longitude",
    y = "Latitude"
  )

p1 <- p +
  facet_wrap(~time)
```

Let's first look at the posterior densities for each time step 

```{r post densities timestep}
p1
```

Next let's look at the densities combined to gather a better understanding of the movement over the 8 timesteps. 

```{r post densities combined}
p
```

Next we can plot the posterior distributions for `alpha0`, `alpha1`, `p0`, and `sigma`. We need to first make `fish` and `time` a `character` and then to make plotting easier we need to make the `data.frame` in long format. 

```{r plot params}
param_draws$fish <- as.character(param_draws$fish)
param_draws$time <- as.character(param_draws$time)

param_draws_long <- param_draws |>
  pivot_longer(
    cols = -c(".chain", ".iteration", ".draw", "lp__", "fish", "time"),
    names_to = "param",
    values_to = "est"
  )
```

We can now plot those posterior distributions of the model parameters.    

```{r plot param}
p_param <- ggplot(
  data = param_draws_long,
  aes(x = time, y = est, fill = fish),
) +
  geom_violin() +
  facet_wrap(~param, scale = "free_y") +
  scale_fill_manual(name = "Fish ID", values = "#2768F5") +
  theme_bw() +
  theme(
    panel.grid = element_blank(),
    strip.background = element_blank()
  ) +
  labs(x = "Time Bin", y = "Estimate")

p_param
```
We can see that they are all quite tightly distributed with `p0` indicating that detection probability at a distance of `0` is between 48 - 54 %, while the `sigma` is around 1 km. 

Lastly, we can plot the predictive posterior check using `{tidybayes}`. First we need to make 
detection counts as a vector from our `build_count()` object. 

```{r y_obs }
y_obs <- as.vector(ps_count_example[!is.na(ps_count_example)])
```
Next we can plot the densities using `ppc_dens_overlay()` from `{bayesplot}`.
```{r ppc dens}
ppc <- ppc_dens_overlay(y = y_obs, yrep = yrep$yrep)
ppc
```

We can see our predictive posterior check distribution for 10 draws closely lines up with 
our observed detection counts indicating that the model fits the data well. 

Congratulations we have successfully run a Bayesian standard point processing/detection probability model for one Lake Trout for 8, 1 hour time bins (8 hrs) in Parry Sound, a large embayment of Georgian Bay, Lake Huron. We can take the results and write up how this individual behaved for a given time period based on the model results!